From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-06-27 draft_v1/draft_v2.tex
update draft_v2

\date{2025-06-24}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between domain-specific
precariousness indicators, and depressive symptoms. Our multi-resolution
analysis---spanning both aggregate and symptom-level
constructs---identified financial stress as a robust driver of
depression, and highlighted key symptoms, such as sleep disturbance and
depressed mood, as likely initiators or reinforcers of downstream
precariousness. To explore how these structural insights shape
intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precariousness under varying levels of financial stress reduction.
Simulation results revealed that the system's responsiveness depends not
only on the intensity of external intervention but also on internal
structural features such as feedback strength and noise. In particular,
some configurations exhibited bistability, requiring interventions to
surpass a critical threshold to trigger lasting improvement, while
others responded more proportionally. Together, these findings emphasize
that internal feedback dynamics---not just external adversity---can
shape the impact of mental health interventions. More broadly, our study
offers a scalable framework for integrating causal discovery and
dynamical simulation to better understand and inform policy responses to
complex psychosocial phenomena.
\end{abstract}

@@

As illustrated in Figure~\ref{fig-concept}, the model includes three
components: depression (\(D\)), represented by the PHQ-9 sum score,
social precariousness (\(P\)), and external financial stress (\(S\)).
Depression and precariousness are modeled as coupled state variables
that influence each other over time, with \(S\) acting as a shared
input. These dynamics are formalized as a pair of nonlinear stochastic
differential equations (SDEs):

\[
\begin{aligned}
dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
\end{aligned}
\]

Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
\(S\) is the external stressor associated with financial stress. The
parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
which variable adjusts toward its input-driven target state. The
coefficients \(\alpha_1, \alpha_2\) and \(\beta_1, \beta_2\) represent
the strengths of the causal influences between variables. Stochastic
fluctuations are introduced through the terms \(\sigma_1 dW_1\) and
\(\sigma_2 dW_2\), where \(dW_1\) and \(dW_2\) are independent Wiener
processes. The system uses the hyperbolic tangent function (\(tanh\)) to
bound the influence of inputs, ensuring smooth, saturating dynamics and
preventing runaway feedback loops. Importantly, the nonlinearity
introduced by the \(tanh\) function also allows the system to exhibit
bistability, supporting the existence of two stable states: one
corresponding to low depression and precariousness, and the other to
high depression and precariousness.

@@

\subsubsection{Model Calibration}\label{model-calibration}

To identify realistic parameter settings for the simulation model, we
calibrate six free parameters---\(\alpha_1\), \(\alpha_2\), \(\beta_1\),
\(\beta_2\), and noise amplitudes \(\sigma_1\), \(\sigma_2\)---using the
\emph{Non-dominated Sorting Genetic Algorithm II} (NSGA-II) (Deb et al.,
2002), a multi-objective evolutionary optimization method well-suited
for noisy and high-dimensional search spaces. The goal is to match
simulated system behavior to empirical summary statistics derived from
the HELIUS dataset. Specifically, we aim to match the observed means and
variances of depression (\(D\)) and social precariousness (\(P\)), along
with all three partial correlations among \(D\), \(P\), and financial
stress (\(S\)): \(\rho_{DP.S}\), \(\rho_{DS.P}\), and \(\rho_{PS.D}\).

Given the stochastic nature of the model, each candidate parameter set
is evaluated by simulating 1,000 virtual individuals, each assigned a
fixed \(S\) value sampled from the empirical distribution. From these
simulations, we compute population-level summary statistics based on the
final states. The fitness of each parameter set is measured by how
closely its summary statistics matched the empirical data, using the
total absolute deviation across all objectives.

To explore optimal parameter configurations, we used the NSGA-II
algorithm as implemented in the \texttt{nsga2R} package (Franz \&
Nakamura, 2015). This algorithm is inspired by natural selection and
evolves a population of candidate solutions over multiple generations to
balance competing objectives. We began with a population of 500 randomly
generated parameter sets and allowed them to evolve over 300
generations. During this process, solutions were selected, combined, and
mutated using standard settings: a crossover probability of 0.9 (which
controls how often solutions are combined), a mutation probability of
0.25 (introducing random variation), and a tournament size of 2 (used to
pick the best candidates for reproduction). The result is a Pareto
front---a set of solutions that are considered optimal in the sense that
no single objective can be improved without worsening another. From this
diverse set of trade-off solutions (see Figure~\ref{fig-pareto}), we
selected the 30 parameter sets with the lowest overall error across
objectives. These top-performing configurations were then used in the
simulation phase to evaluate the potential effects of different
interventions.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/paretoplot.png}

@@

\caption{\label{fig-pareto}Pareto front of simulated parameter sets.
\emph{Each point represents a solution evaluated by its total variance
error (x-axis) and total partial correlation error (y-axis), with color
indicating total mean error. Red crosses highlight the top 30 solutions
used for simulation.}}

@@

To evaluate the effects of hypothetical support, we simulate scenarios
in which each individual's financial stress level (\(S\)) is
proportionally reduced toward the minimum observed value. This models a
population-wide intervention applied at varying strengths --- from no
reduction in stress (0\%) to complete alignment with the minimum
observed stress level (100\%).

For each of the top 30 parameter sets identified through calibration, we
run simulations on 300 virtual individuals across this full range of
intervention levels. In each simulation, we observe how depression
(\(D\)) and precariousness (\(P\)) evolve over time, capturing their
final values after the system stabilizes. This setup allows us to
evaluate how different internal model dynamics (e.g., feedback strength)
influence the system's response to external change.

@@

To evaluate how internal system structure influences the effect of
external support, we simulated outcomes under a range of intervention
strengths across the top 30 parameter sets from the calibrated model
ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
levels of depression (\(D\)) and precariousness (\(P\)) after
simulation, with each line representing one parameter set. Lines are
color-coded by combined feedback and input strength, calculated as the
product \(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).

Parameter sets with stronger interaction dynamics exhibited more
pronounced reductions in both depression and precariousness following
intervention. At full intervention---modeled as shifting the external
stressor (\(S\)) to its observed minimum---the reduction in depression
ranged from {[}0.68\emph{SD} -- 0.97\emph{SD}{]}, and for social
precariousness from {[}0.77\emph{SD} -- 0.94\emph{SD}{]}, corresponding
to the lowest to highest values of parameter strength.\footnote{Standard
  deviations (\emph{SD}) are based on the original scale of each
  variable: \(\textit{SD}_D = 1.0\) for depression (\(D\)) and
  \(\textit{SD}_P = 0.9\) for precariousness (\(P\)). Parameter strength
  is defined as the product
  \(\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2\), reflecting
  the combined influence of internal feedback and external input
  sensitivity.} These results suggest that systems with stronger
internal dynamics are more responsive to external change, enabling
targeted interventions to more effectively shift the system toward a
lower-risk equilibrium.

To better understand these dynamics, we focused on two illustrative
parameter sets: one with strong feedback/input coupling and low noise (a
\emph{bistable} candidate), and one with weaker coupling and higher
noise (a \emph{unistable} candidate). In nonlinear systems, bistability
refers to the existence of two distinct stable equilibria---meaning the
system's long-term outcome depends on its starting point or
perturbations.

In our model, bistability arises from the interaction between strong
internal coupling---particularly the product
\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
\(\tanh(\cdot)\) activation function saturates at high input values,
meaning sufficient gain is required to push the system into nonlinear
regimes where multiple stable points emerge. If the feedback gain is too
low, or the system is overwhelmed by noise, it behaves more linearly and
settles into a single equilibrium---yielding a unistable
configuration.\footnote{This is similar to a well-known pattern in
  dynamical systems called a \emph{pitchfork bifurcation}. In simple
  terms, when the system's internal feedback becomes strong enough---for
  example, when \(k\) in the equation \(x_{t+1} = \tanh(k x_t)\) passes
  a certain threshold---the system changes from having one stable state
  to having two possible stable states (and one unstable one in
  between). In our model, this kind of shift happens when the feedback
  loop between depression and precariousness---captured by the product
  \(\alpha_2 \times \beta_2\)---becomes strong. It allows the system to
  settle into different long-term states, depending on initial
  conditions.}

The bottom panel of Figure~\ref{fig-intervention} illustrates these
contrasting behaviors. In the bistable model, a clear bifurcation
emerges: small interventions have little effect, but once a threshold is
crossed, the system abruptly transitions to a low-risk state. In the
unistable model, the transition is gradual and continuous, with outcomes
improving incrementally as stress is reduced.

These findings underscore that the impact of intervention depends not
only on its intensity, but on the system's internal configuration. In
bistable regimes, small or moderate interventions may be insufficient
unless they push the system past a tipping point. Unistable systems, by
contrast, respond more linearly, offering greater predictability and
flexibility in designing incremental interventions.

This distinction is especially relevant for mental health interventions
in populations exposed to persistent stressors like housing instability,
financial strain, or job insecurity. In such contexts, external
adversity may be reinforced through internal feedback loops between
psychological distress and precarious conditions. Recognizing when a
system is bistable may help identify when strong, early, or multifaceted
interventions are needed to produce lasting change---while unistable
dynamics may justify more gradual or targeted efforts.

Tailoring intervention strategies to the system's underlying structure
may enhance their effectiveness and sustainability, particularly in
addressing the compounding effects of socioeconomic precariousness on
mental health.

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/intervention_plot2.pdf}

@@

\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
precariousness (\(P\)) across varying intervention strengths, simulated
over the top 30 parameter sets. Each line represents one parameter set
and is colored by its combined feedback and input strength.}
\textbf{Bottom}: Smoothed Density Heatmap of Final \(D\) and \(P\) for
Two Parameter Sets. \emph{One exhibits unistability (top two panels) and
the other exhibits bistability (bottom two panels). Bistable dynamics
reveal a sharp transition between distinct outcome regimes, while
unistable dynamics show gradual, continuous change across the
intervention gradient.}}

@@

To examine how such structures shape responsiveness to support, we
developed a simplified two-variable nonlinear dynamical model capturing
the feedback between depression and precariousness, modulated by
financial stress. Calibrated against empirical summary statistics from
the HELIUS dataset, the model allowed us to simulate how these dynamics
evolve under varying levels of stress reduction. The simulations
revealed that intervention effectiveness depends not just on the
magnitude of external stressors, but on the internal configuration of
the system itself. In bistable regimes---characterized by strong
feedback and low noise---the system could remain stuck in a high-risk
state unless an intervention crosses a critical threshold, tipping it
into a healthier equilibrium. By contrast, unistable systems---driven by
weaker coupling or greater noise---responded more gradually and
proportionally to increasing support. These dynamics carry practical
policy implications. In bistable systems, even moderate interventions
may have little effect unless they are forceful enough to tip the system
toward a lower-risk state. This underscores the importance of early or
intensive support when targeting self-reinforcing cycles of depression
and precariousness. Unistable systems, by contrast, are more responsive
to incremental support, suggesting that sustained, moderate
interventions may be sufficient in such contexts.

@@

Finally, the structure of the dynamical model introduces its own
limitations. While the model was designed to reflect key features of the
discovered causal structure, it represents a highly simplified system:
only two variables---depression and social precariousness---modulated by
external financial stress. This abstraction allowed for tractable
analysis and visual intuition, particularly around nonlinear behaviors
such as bistability, but omits other potentially influential constructs
highlighted in the causal graphs, including relational stress,
employment conditions and housing conditions. Moreover, the model's
nonlinearity---introduced via the hyperbolic tangent function---was
chosen for convenience rather than empirical justification. While this
function supports saturation and the emergence of tipping points, the
presence of bistability is a product of simulation, not empirical
verification. It remains possible that simpler, linear dynamics could
account for the observed data just as well. Future work should further
investigate whether bistability is a necessary feature of
depression--precariousness dynamics, and explore alternative functional
forms---both linear and nonlinear---using richer data and stronger
empirical constraints.

Our results reveal that the effectiveness of interventions targeting
depression and precariousness is shaped not only by the strength of
external stressors but also by the internal structure of the system. The
interplay between feedback strength and noise determines whether a
system responds gradually or only after crossing a tipping
point---highlighting the need for intervention strategies attuned to
these underlying dynamics.

These findings underscore the importance of considering internal
feedback---not just external adversity---when designing interventions.
Mental health systems are not always passively responsive to support;
their structure can constrain how much impact any given policy has
(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By diagnosing these
structural features---through tools like causal discovery and
computational modeling---researchers and policymakers can better tailor
interventions to the underlying mechanics of vulnerability.

Beyond these specific findings, this study offers a generalizable
framework for studying complex mental health phenomena. Causal discovery
enables principled hypothesis generation from observational data, while
dynamical modeling helps test how those structures behave under
intervention. Together, these methods move us from descriptive patterns
to mechanistic insight, with practical relevance for real-world mental
health policy.

Future work could build richer dynamical models grounded in
higher-resolution empirical data---potentially incorporating time,
interventions, or more granular psychosocial constructs. Such extensions
would help validate whether mechanisms proposed here, such as
bistability, manifest in real-world trajectories. Ultimately, we hope
this study encourages further integration of causal structure learning
and system-level modeling to improve mental health outcomes under
conditions of socioeconomic precariousness.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-deb2002fast}
Deb, K., Pratap, A., Agarwal, S., \& Meyarivan, T. (2002). A fast and
elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
Transactions on Evolutionary Computation}, \emph{6}(2), 182--197.

@@

@@

@@

@@

@@

>>>>

\date{2025-06-27}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates how the internal structure of
depression--precariousness dynamics shapes responsiveness to external
support. Using cross-sectional data from the HELIUS study, we first
apply cycle-capable causal discovery algorithms to estimate directional
dependencies between financial stress, domain-specific precariousness,
and depressive symptoms. Our multi-resolution analysis---spanning both
aggregate and symptom-level constructs---highlights financial stress as
a plausible upstream driver of depression, with symptoms such as sleep
disturbance and depressed mood emerging as initiators of broader social
vulnerability. To translate these structural insights into dynamic
predictions, we construct a family of analytically calibrated linear
stochastic models. Simulations reveal that variations in internal
structure can substantially alter system responses to financial stress
reduction. Some models show rapid but transient improvements in
depression, while others adjust more slowly yet sustain gains over time.
Precariousness tends to respond more immediately across all
configurations, though its trajectory also shifts subtly with structural
variation. Together, these findings demonstrate that intervention
outcomes depend not only on the magnitude of external input but also on
latent properties of the system. More broadly, this study illustrates
how combining causal discovery with dynamical modeling supports
principled, simulation-based exploration of intervention
scenarios---grounded in data, yet able to probe counterfactual
structure. This approach offers a scalable framework for diagnosing
hidden system constraints and informing targeted public health
strategies in complex psychosocial domains.
\end{abstract}

@@

As illustrated in Figure~\ref{fig-concept}, the model includes three
components: depression (\(D\)), represented by the PHQ-9 sum score,
social precariousness (\(P\)), and external financial stress (\(S\)).
Depression and precariousness are modeled as coupled state variables
that influence each other over time, with \(S\) acting as a shared
input. These dynamics are formalized as a pair of linear stochastic
differential equations (SDEs):

\[
\begin{aligned}
dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
\end{aligned}
\]

Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
\(S\) is the external stressor associated with financial stress. The
parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
which each variable adjusts toward its input-driven target state. The
coefficients \(\alpha_{PS}\) and \(\alpha_{DS}\) quantify the direct
effects of the stressor on \(P\) and \(D\), while \(\alpha_{DP}\) and
\(\alpha_{PD}\) represent the cross-influences between depression and
precariousness. Stochastic fluctuations are introduced through the noise
terms \(\sigma_P, dW_P\) and \(\sigma_D, dW_D\), with \(dW_P\) and
\(dW_D\) representing independent Wiener processes. This linear
structure offers a transparent and analytically tractable framework for
examining how variations in external financial stress influence the
population-level dynamics of depression and social precariousness.
Crucially, the model's simplicity enables direct calibration from
empirical data: all parameters can be derived analytically from observed
variances and covariances in the HELIUS cohort. The following
Section~\ref{sec-calibration} details the procedure used to perform this
calibration.

@@

\subsubsection{Model Calibration}\label{sec-calibration}

In our linear model, all parameters are derived analytically to
reproduce the empirical second-order moments observed in the HELIUS
dataset. The system treats depression (\(D\)) and social precariousness
(\(P\)) as jointly evolving under the influence of a shared external
stressor (\(S\)), forming a coupled system of Ornstein--Uhlenbeck
processes. Under the assumption of stationarity, the steady-state
covariances between these variables admit closed-form solutions for the
system's structural parameters.

We fix the adjustment rates \(\lambda_D = \lambda_P = 1\), which
normalizes the time scale given the cross-sectional nature of the HELIUS
data. To resolve the remaining degrees of freedom, we treat the feedback
from precariousness to depression (\(\alpha_{DP}\)) as a free parameter,
and solve algebraically for all other coefficients --- including the
direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
feedback from depression to precariousness (\(\alpha_{PD}\)), and the
noise terms (\(\sigma_D, \sigma_P\)). These solutions are expressed as
closed-form functions of the empirical covariances: \(\mathrm{Var}(D)\),
\(\mathrm{Var}(P)\), \(\mathrm{Var}(S)\), \(\mathrm{Cov}(D, P)\),
\(\mathrm{Cov}(D, S)\), and \(\mathrm{Cov}(P, S)\).

This structure creates a family of analytically calibrated models, each
consistent with the same observed covariances but differing in their
internal configuration. However, we constrain the space of valid
solutions by imposing two empirical and theoretical conditions: (1) all
structural parameters must be positive, reflecting the assumption that
stress and precariousness contribute adversely to depression; and (2)
the system must be dynamically stable, ensuring convergence to a
stationary distribution. These constraints restrict the allowable range
of \(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
Section~\ref{sec-analcalibration}).

Figure~\ref{fig-tradeoff} illustrates how the remaining parameters
adjust as \(\alpha_{DP}\) increases to maintain consistency with the
empirical covariances. As \(\alpha_{DP}\) increases, \(\alpha_{PD}\)
must decline to preserve stability, since the product
\(\alpha_{DP} \cdot \alpha_{PD}\) determines the strength of the
feedback loop and must remain below 1. This creates a compensatory
trade-off: as influence from \(P\) to \(D\) strengthens, the influence
from \(D\) to \(P\) weakens. Meanwhile, the external coupling terms ---
\(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to absorb variance in
a way that maintains alignment with the empirical data. Notably, the
noise amplitudes \(\sigma_D\) and \(\sigma_P\) decline slightly as
\(\alpha_{DP}\) increases. This indicates that more of the observed
variability is captured by deterministic interactions, with less left to
be explained by stochastic fluctuations.

Rather than being a limitation, this flexibility enables exploration of
a spectrum of equally data-consistent feedback regimes. Each represents
a different plausible causal architecture, offering insight into how
internal dynamics modulate the system's responsiveness to external
stress --- even when the observed covariances remain unchanged.

@@

\includegraphics[width=1\textwidth,height=\textheight]{img/param_tradeoff.pdf}

@@

\caption{\label{fig-tradeoff}Trade-off curves showing how calibrated
parameters vary with \(\alpha_{DP}\). \emph{Each line shows how a
different parameter adjusts in response to changes in \(\alpha_{DP}\) to
maintain consistency with empirical covariances from the HELIUS data.
All configurations shown are equally consistent with the data,
illustrating how internal dynamics adjust as feedback strength
changes.}}

@@

To evaluate how internal structure affects system responsiveness, we
simulate a family of linear models subjected to external intervention.
Each model corresponds to a different value of \(\alpha_{DP}\),
representing the strength of influence from social precariousness to
depression. The range spans from 0 (no coupling) to 0.65 (maximal
admissible value), with all other parameters derived analytically to
preserve the empirical covariances observed in the HELIUS data (see
Figure~\ref{fig-tradeoff}).

In each simulation, financial stress (\(S\)) is proportionally reduced
toward the lowest observed value, modeling a population-wide
intervention applied at varying intensities (0--100\%). For each model,
we simulate 1000 virtual individuals whose baseline stress levels are
sampled from the empirical distribution and scaled according to the
intervention level.

We assess the system's behavior using two complementary approaches:

\begin{itemize}
\item
  Snapshot analysis captures the average levels of depression (\(D\))
  and precariousness (\(P\)) at selected time points (\(t = 100\),
  \(300\), \(500\)) following intervention onset. Since all linear
  models converge to the same equilibrium, this highlights how internal
  feedback affects the speed and pathway of adjustment.
\item
  Full trajectory simulation tracks system dynamics over time for two
  representative models: one with minimal (\(\alpha_{DP} = 0\)) and one
  with maximal (\(\alpha_{DP} = 0.65\)) feedback. Each simulation
  includes an intervention period followed by its withdrawal, revealing
  differences in reactivity, delay, and recovery stability.
\end{itemize}

Together, these simulations demonstrate how internal structure governs
not just whether change occurs, but how it unfolds --- shaping the pace,
persistence, and variability of response to external support.

@@

To assess how internal structure influences the system's responsiveness
to external support, we simulate a set of linear models that differ in
the assumed strength of influence from precariousness to depression,
\(\alpha_{DP}\). For each model, we generate a virtual population of
1,000 individuals and examine system behavior in two complementary ways:
by analyzing average outcomes at selected time points following
intervention onset, and by simulating full temporal trajectories for two
representative cases at the extremes of the \(\alpha_{DP}\) range.

Figure~\ref{fig-intervention} summarizes how the internal structure of
the model---specifically, variation in \(\alpha_{DP}\)---shapes the
system's short-term and long-term responses to external support. Panel A
shows the mean levels of depression (\(D\)) and precariousness (\(P\))
at time steps 100, 300, and 500 after intervention onset, across a range
of \(\alpha_{DP}\) values. Each line represents a different model,
color-coded from low (purple) to high (yellow) values of
\(\alpha_{DP}\).

In the early phase of intervention (\(t = 100\)), depression and
precariousness exhibit contrasting patterns. Depression declines more
rapidly in models with lower \(\alpha_{DP}\), where its dynamics are
driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
increases, \(D\) becomes more dependent on precariousness (\(P\)), which
itself adjusts more slowly---delaying the depressive response.
Precariousness, by contrast, shows slightly faster early reductions in
models with higher \(\alpha_{DP}\). This reflects a shift in influence:
as \(\alpha_{DP}\) increases, \(P\) becomes more directly responsive to
\(S\) (via increasing \(\alpha_{PS}\)) and less influenced by \(D\),
allowing it to adjust more promptly to changes in the stressor.

Despite these short-term differences, all models ultimately converge to
similar steady-state levels under sustained intervention. By time step
300, convergence is already substantial---though some variation remains,
particularly in \(P\), where gradients between models are still visible.
By step 500, trajectories have nearly fully aligned across all
configurations, consistent with the shared equilibrium structure of the
linear system. This convergence highlights that the key distinctions lie
not in whether intervention is effective, but in how quickly and through
which pathways change unfolds.

Panel B shows full time trajectories for two representative models: one
with no coupling (\(\alpha_{DP} = 0\)) and one at the upper admissible
limit (\(\alpha_{DP} = 0.65\)). In both, intervention begins at step 300
and ends at step 1500. In the uncoupled model, depression responds
rapidly and rebounds shortly after support is removed. In the
high-\(\alpha_{DP}\) model, \(D\) responds more gradually and recovers
more slowly---reflecting its greater dependence on the slower-changing
\(P\). Precariousness declines in both models during intervention and
returns to baseline afterward, but with a slightly sharper drop in the
higher \(\alpha_{DP}\) condition. This is because \(P\) becomes more
directly sensitive to \(S\) as \(\alpha_{PS}\) increases with
\(\alpha_{DP}\).

These asymmetries further highlight how internal structure---not just
external inputs---shapes the system's trajectory. Specifically, changes
in coupling parameters reconfigure how stress reduction propagates
through the system, altering the speed and pathway of response even when
long-run outcomes remain unchanged.

@@

\includegraphics{img/intervention_linear.pdf}

@@

\caption{\label{fig-intervention}System response to external
intervention across variations in internal structure.
\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
precariousness (\(P\)) at time steps 100, 300, and 500, across different
values of \(\alpha_{DP}\) and intervention strengths. Models with lower
\(\alpha_{DP}\) show faster early reductions in \(D\), while higher
\(\alpha_{DP}\) leads to more immediate declines in \(P\). \textbf{Panel
B}: Full time trajectories for two models---one with no coupling
(\(\alpha_{DP} = 0\)) and one with strong coupling
(\(\alpha_{DP} = 0.65\))---illustrating system response and recovery
following intervention onset (step 300) and withdrawal (step 1500).
Depression responds more gradually and persistently in the high
\(\alpha_{DP}\) model, whereas precariousness adjusts more quickly.}}

@@

To explore how internal structure shapes response to external support,
we developed a linear stochastic model of depression and precariousness,
influenced by financial stress. Crucially, we analytically calibrated
this model to match empirical covariances from the HELIUS data, ensuring
that all parameter sets were observationally equivalent but differed in
their internal configuration. We treated the parameter
\(\alpha_{DP}\)---the effect of precariousness on depression---as a free
variable, and derived all other parameters accordingly, yielding a
family of valid models consistent with the data. Simulation results
showed that even with identical steady-state covariances, differences in
internal structure produced systematic variation in the timing and
profile of system responses to intervention. In models with lower
\(\alpha_{DP}\), depression responded more quickly to reduced stress
because its dynamics were more directly coupled to the external input.
In contrast, as \(\alpha_{DP}\) increased and depression became more
indirectly influenced---mediated by the slower-changing precariousness
variable---its response was more gradual and persistent. Precariousness,
while generally more directly responsive across all models, also showed
subtle differences, with slightly faster early declines in
configurations where it was more sensitive to external stress. These
results demonstrate that intervention impact depends not only on the
external conditions (e.g., stress reduction), but also on the internal
structure of the system---how variables are coupled and through which
pathways change flows. Tailoring interventions to system
configuration---using more intensive or earlier support in systems with
indirect or delayed pathways, and sustained support in systems with
rapid but unstable responses---may help optimize mental health outcomes
under conditions of social and economic adversity.

@@

Finally, the structure of our dynamical model introduces important
limitations. While the linear formulation offered analytical
tractability and transparent calibration, it likely oversimplifies the
true dynamics of depression and socioeconomic precariousness. A growing
body of evidence suggests that psychosocial systems are fundamentally
nonlinear. Empirical and theoretical work points to phenomena such as
threshold effects, where improvements only occur after a tipping point
is crossed (Leemput et al., 2014; Park et al., 2025); saturation, where
the impact of support diminishes as basic needs are met (Cramer et al.,
2010); and path dependence, where past adversity shapes sensitivity to
future stressors (Wichers et al., 2016). These behaviors are
fundamentally incompatible with linear dynamics, yet they may be
essential for understanding why some individuals remain trapped in
high-risk states while others recover more easily. Future modeling
efforts should therefore incorporate mechanisms such as saturating
input-response functions (e.g., sigmoidal or tanh forms) and
state-dependent feedback (where coupling strength varies by system
state). Incorporating such features would allow models to capture not
just average effects, but also the critical conditions under which
meaningful change becomes possible---clarifying whether some subgroups
require forceful, threshold-crossing interventions, while others may
benefit from sustained, moderate support. Ultimately, embracing
nonlinear dynamics may be key to understanding whether different
populations face genuinely distinct stability regimes---and whether
effective intervention requires shifting the system across a critical
boundary, rather than simply nudging it along its current trajectory.

Our findings demonstrate that intervention effectiveness depends not
only on external stressors, but also on the internal structure of the
system. While all models converge to the same long-term outcome,
internal coupling shapes how quickly and through which pathways change
unfolds. This highlights the importance of designing interventions that
are sensitive to the system's internal architecture.

As public systems research has long shown, social and psychological
systems do not always respond passively to external inputs; internal
constraints can limit the impact of even well-intentioned policies
(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By using causal
discovery and computational modeling, researchers and policymakers can
better identify these structural constraints and tailor interventions
accordingly.

More broadly, this study offers a generalizable framework for
investigating complex mental health dynamics. Causal discovery supports
principled hypothesis generation from observational data, while
dynamical modeling enables exploration of how hypothesized structures
behave under external change. Together, these tools move us beyond
static descriptions toward mechanistic insight---with direct relevance
for real-world mental health policy and intervention design.

Future research should build on this foundation by developing richer,
more nuanced dynamical models grounded in longitudinal and
interventional data. Incorporating time, multiple domains of precarity,
and adaptive feedback processes would help assess whether the mechanisms
proposed here---such as delayed response, saturation, or
multi-stability---manifest in actual mental health trajectories.
Ultimately, we hope this study encourages further integration of causal
structure learning and systems modeling in efforts to understand and
improve mental health outcomes under conditions of socioeconomic
precariousness.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-butterworth2012role}
Butterworth, P., Olesen, S. C., \& Leach, L. S. (2012). The role of
hardship in the association between socio-economic position and
depression. \emph{Australian \& New Zealand Journal of Psychiatry},
\emph{46}(4), 364--373.

\bibitem[\citeproctext]{ref-cramer2010comorbidity}
Cramer, A. O., Waldorp, L. J., Van Der Maas, H. L., \& Borsboom, D.
(2010). Comorbidity: A network perspective. \emph{Behavioral and Brain
Sciences}, \emph{33}(2-3), 137--150.

\bibitem[\citeproctext]{ref-deb2002fast}
Deb, K., Pratap, A., Agarwal, S., \& Meyarivan, T. (2002). A fast and
elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
Transactions on Evolutionary Computation}, \emph{6}(2), 182--197.

@@

\bibitem[\citeproctext]{ref-van2014critical}
Leemput, I. A. van de, Wichers, M., Cramer, A. O., Borsboom, D.,
Tuerlinckx, F., Kuppens, P., Van Nes, E. H., Viechtbauer, W., Giltay, E.
J., Aggen, S. H., et al. (2014). Critical slowing down as early warning
for the onset and termination of depression. \emph{Proceedings of the
National Academy of Sciences}, \emph{111}(1), 87--92.

@@

\bibitem[\citeproctext]{ref-lorant2003socioeconomic}
Lorant, V., Deliège, D., Eaton, W., Robert, A., Philippot, P., \&
Ansseau, M. (2003). Socioeconomic inequalities in depression: A
meta-analysis. \emph{American Journal of Epidemiology}, \emph{157}(2),
98--112.

@@

\bibitem[\citeproctext]{ref-ridley2020poverty}
Ridley, M., Rao, G., Schilbach, F., \& Patel, V. (2020). Poverty,
depression, and anxiety: Causal evidence and mechanisms. \emph{Science},
\emph{370}(6522), eaay0214.

@@

\bibitem[\citeproctext]{ref-wichers2016critical}
Wichers, M., Groot, P. C., Psychosystems, E., Group, E., et al. (2016).
Critical slowing down as a personalized early warning signal for
depression. \emph{Psychotherapy and Psychosomatics}, \emph{85}(2),
114--116.

@@

\subsection{Analytical Calibration of the Linear Feedback
Model}\label{sec-analcalibration}

We consider a continuous-time, linear stochastic differential equation
(SDE) model describing the coupled dynamics between depression (\(D\))
and social precariousness (\(P\)), each influenced by an external
stressor (\(S\)). The system is specified as:

\[
\begin{aligned}
\frac{dD}{dt} &= \alpha_{DS} S + \alpha_{DP} P - D + \sigma_D \cdot \eta_D(t), \\
\frac{dP}{dt} &= \alpha_{PS} S + \alpha_{PD} D - P + \sigma_P \cdot \eta_P(t),
\end{aligned}
\]

where:

\begin{itemize}
\tightlist
\item
  \(S\) is a fixed or exogenous input (external financial stress),
\item
  \(\alpha_{DP}\) represents the feedback from precariousness to
  depression,
\item
  \(\alpha_{PD}\) represents the feedback from depression to
  precariousness,
\item
  \(\alpha_{DS}\) and \(alpha_{PS}\) capture the direct effects of
  \(S\),
\item
  \(\sigma_D, \sigma_P\) are stochastic noise amplitudes,
\item
  \(\eta_D(t), \eta_P(t)\) are independent standard Wiener processes.
\end{itemize}

We fix the decay rates \(\lambda_D = \lambda_P = 1\), normalizing the
time scale due to lack of temporal resolution in the HELIUS dataset.

\subsubsection{Stationarity Assumption}\label{stationarity-assumption}

We assume the joint distribution of \((D, P)\) reaches stationarity,
leading to a multivariate Ornstein--Uhlenbeck process. Under this
assumption, the system has a unique stationary distribution
characterized by its covariance structure, which we match to the
empirical data.

Let:

\[
\Sigma_{DD} = \mathrm{Var}(D), \quad \Sigma_{PP} = \mathrm{Var}(P), \quad \Sigma_{SS} = \mathrm{Var}(S),
\]

\[
\Sigma_{DP} = \mathrm{Cov}(D, P), \quad \Sigma_{DS} = \mathrm{Cov}(D, S), \quad \Sigma_{PS} = \mathrm{Cov}(P, S).
\]

From the HELIUS data, we use the following observed values:

\(\Sigma_{DD} = 1.000, \quad \Sigma_{PP} = 0.807, \quad \Sigma_{SS} = 0.743, \quad \Sigma_{DP} = 0.374, \quad \Sigma_{DS} = 0.357, \quad \Sigma_{PS} = 0.256.\)

\subsubsection{Analytical Parameter
Derivation}\label{analytical-parameter-derivation}

We consider a linear stochastic dynamical system in which depression
(\(D\)) and social precariousness (\(P\)) evolve under mutual influence
and are jointly affected by an exogenous stressor (\(S\)). We assume
stationary dynamics governed by a two-dimensional Ornstein--Uhlenbeck
process:
